Causal Bandits: Learning Good Interventions via Causal Inference

### 2016-05-01 causal-bandit/nips_2016/introduction.tex
adds typo corrections to first 3 sections

Medical drug testing, policy setting, and other scientific processes are commonly framed and analysed in the language of sequential experimental design and, in special cases, as bandit problems~\citep{Robbins1952,Chernoff1959}. 
In this framework, single actions (\eg, experiments or interventions) from a pre-determined set are repeatedly performed in 
order to evaluate their effectiveness via feedback from a single, real-valued reward signal.
We propose a generalisation of the standard model by assuming that, in additional to the reward signal, the learner observes the values of a number of covariates 
drawn from a probabilistic causal model~\citep{Pearl2000}.
Causal models are commonly used in disciplines where explicit experimentation may be difficult such as social science, demography and economics.
For example, when predicting the effect of changes to childcare subsidies on workforce participation, or school choice on grades. 
Results from causal inference relate observational distributions to interventional ones, allowing the outcome of an intervention to be predicted without
explicitly performing it.
By exploiting the causal information we show, theoretically and empirically, how non-interventional observations can be used to improve the rate at 
which high-reward actions can be identified.

The type of problem we are concerned with is best illustrated with an example. 
Consider a farmer wishing to optimise the yield of her crop. 
She knows that crop yield is only affected by temperature, a particular soil nutrient, and moisture level but the precise effect of their combination is unknown.
In each season the farmer has enough time and money to intervene and control at most one of these variables:
deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed a through a choice of fertilizer; and irrigation or rain-proof covers will keep the soil wet or dry.
When not intervened upon, the temperature, soil, and moisture vary naturally from season to season due to weather conditions and these are all observed along with the final crop yield at the end of each season.
How might the farmer best experiment to identify the single, highest yielding intervention in a limited number of seasons?
%without sacrificing too much crop yield (relative to always choosing the best intervention) in the process?

@@

As alluded to above, causal bandit problems with $N$ possible interventions can be treated as a classical $N$-armed bandit problem by simply ignoring the causal model and extra observations that are revealed each round.
In the farming example, the farmer could treat the six possible interventions as arms, ignore the observations of the non-intervened variables, and apply an existing best-arm identificaion algorithm with well understood simple regret guarantees \citep{Jamieson2013}.
However, as we show in \S\ref{sec:simple-regret}, this approach ignores the extra information available in the non-intervened variables and subsequently yields sub-optimal performance.

>>>>

Medical drug testing, policy setting, and other scientific processes are commonly framed and analysed in the language of sequential experimental design and, in special cases, as bandit problems~\citep{Robbins1952,Chernoff1959}. 
In this framework, single actions (\eg, experiments or interventions) from a pre-determined set are repeatedly performed in 
order to evaluate their effectiveness via feedback from a single, real-valued reward signal.
We propose a generalisation of the standard model by assuming that, in addition to the reward signal, the learner observes the values of a number of covariates 
drawn from a probabilistic causal model~\citep{Pearl2000}.
Causal models are commonly used in disciplines where explicit experimentation may be difficult such as social science, demography and economics.
For example, when predicting the effect of changes to childcare subsidies on workforce participation, or school choice on grades. 
Results from causal inference relate observational distributions to interventional ones, allowing the outcome of an intervention to be predicted without
explicitly performing it.
By exploiting the causal information we show, theoretically and empirically, how non-interventional observations can be used to improve the rate at 
which high-reward actions can be identified.

The type of problem we are concerned with is best illustrated with an example. 
Consider a farmer wishing to optimise the yield of her crop. 
She knows that crop yield is only affected by temperature, a particular soil nutrient, and moisture level but the precise effect of their combination is unknown.
In each season the farmer has enough time and money to intervene and control at most one of these variables:
deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed through a choice of fertilizer; and irrigation or rain-proof covers will keep the soil wet or dry.
When not intervened upon, the temperature, soil, and moisture vary naturally from season to season due to weather conditions and these are all observed along with the final crop yield at the end of each season.
How might the farmer best experiment to identify the single, highest yielding intervention in a limited number of seasons?
%without sacrificing too much crop yield (relative to always choosing the best intervention) in the process?

@@

As alluded to above, causal bandit problems with $N$ possible interventions can be treated as a classical $N$-armed bandit problem by simply ignoring the causal model and extra observations that are revealed each round.
In the farming example, the farmer could treat the six possible interventions as arms, ignore the observations of the non-intervened variables and apply an existing best-arm identificaion algorithm with well understood simple regret guarantees \citep{Jamieson2013}.
However, as we show in \S\ref{sec:simple-regret}, this approach ignores the extra information available in the non-intervened variables and subsequently yields sub-optimal performance.

### 2016-05-01 causal-bandit/nips_2016/setup.tex
adds typo corrections to first 3 sections

We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one 
edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some 
arbitrary but unknown, real-valued function $r$. The set of allowed interventions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
Conversely, any causal bandit problem can reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by 
treating each possible intervention as an arm and ignoring all sampled values for the observed variables except for the reward.
Intuitively though, one would expect to perform better by making use of the extra observations.

>>>>

We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one 
edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some 
arbitrary but unknown, real-valued function $r$. The set of allowed interventions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
Conversely, any causal bandit problem can be reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by 
treating each possible intervention as an arm and ignoring all sampled values for the observed variables except for the reward.
Intuitively though, one would expect to perform better by making use of the extra observations.

### 2016-05-01 causal-bandit/nips_2016/simple-regret.tex
adds typo corrections to first 3 sections

The algorithm operates as follows. For the first $T/2$ rounds it chooses $do()$ to collect observational data.
The fact that only link from each $X_1,\ldots,X_N$ to $Y$ is a direct, causal one means that $\P{Y|do(X_i=j)}=\P{Y|X_i=j}$. Thus we can create good estimators of the returns of those actions $do(X_i = j)$ for which $\P{X_i = j}$ is large. 
These are the interventions that are easily estimated ``passively'' (\ie, without playing them explicitly). 
Those actions where $\P{X_i = j}$ is small may not be observed (often) in the first $T/2$ rounds, so
the estimates of the returns of these actions from the first $T/2$ samples could be quite poor. For the remaining $T/2$ rounds
the algorithm makes interventions on the actions $a = do(X_i = j)$ for which $\P{X_i = j}$ is small. The precise allocation is
carefully chosen to minimise the worst-case simple regret. 
The difficulty of the problem depends on $\vec{q}$ and, in particular, how many of the variables have an infrequently occurring value (\ie, small $q_i$ or $1-q_i$). 
Let $I_1 = \emptyset$ and for $\tau \in [0,1)$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} \leq \tau}$.
Define
\eqn{
\label{eq:m-simple}
m(\vec{q}) = \max \set{ d : |I_{\frac{1}{d}}| \leq d}\,.
% m(\vec{q}) = \max \set{i : \sum_{k=1}^N \ind{\min\set{q_k, \,1 - q_k} \leq 1/i} \leq i}\,.
}
This captures a trade off between identifying the low probability variables and not having too many of them.
When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 1$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$, the number of variables.

@@

\begin{theorem}\label{thm:lower}
For all $T$, $\vec{q}$ and all strategies there exists a reward function such that
\eq{
\simpleregret 
\in \Omega\left(\sqrt{\frac{m(\vec{q})}{T}}\right)\,.
}
\end{theorem}

>>>>

The algorithm operates as follows. For the first $T/2$ rounds it chooses $do()$ to collect observational data.
The fact that the only link from each $X_1,\ldots,X_N$ to $Y$ is a direct, causal one means that $\P{Y|do(X_i=j)}=\P{Y|X_i=j}$. Thus we can create good estimators of the returns of those actions $do(X_i = j)$ for which $\P{X_i = j}$ is large. 
These are the interventions that are easily estimated ``passively'' (\ie, without playing them explicitly). 
Those actions where $\P{X_i = j}$ is small may not be observed (often) in the first $T/2$ rounds, so
the estimates of the returns of these actions from the first $T/2$ samples could be quite poor. For the remaining $T/2$ rounds
the algorithm makes interventions on the actions $a = do(X_i = j)$ for which $\P{X_i = j}$ is small. The precise allocation is
carefully chosen to minimise the worst-case simple regret. 
The difficulty of the problem depends on $\vec{q}$ and, in particular, how many of the variables have an infrequently occurring value (\ie, small $q_i$ or $1-q_i$). 
Let $I_1 = \emptyset$ and for $\tau \in [0,1)$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} \leq \tau}$.
Define
\eqn{
\label{eq:m-simple}
m(\vec{q}) = \max \set{ d : |I_{\frac{1}{d}}| \leq d}\,.
% m(\vec{q}) = \max \set{i : \sum_{k=1}^N \ind{\min\set{q_k, \,1 - q_k} \leq 1/i} \leq i}\,.
}
This captures a trade off between identifying the low probability variables and not having too many of them.
When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 1$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$, the number of variables.

@@

\begin{theorem}\label{thm:lower}
For all $T$, $\vec{q}$ and all strategies, there exists a reward function such that
\eq{
\simpleregret 
\in \Omega\left(\sqrt{\frac{m(\vec{q})}{T}}\right)\,.
}
\end{theorem}
